---
title: "NativeLand"
author: "Brian Yandell"
date: "2024-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = ".")
```

This began as a 
[Python to R Conversion](https://www.codeconvert.ai/python-to-r-converter)
from a student project.
See
[native-land.ca API](https://native-land.ca/resources/api-docs/)
for source material.
See also 
[Simple feature geometry](https://tmieno2.github.io/R-as-GIS-for-Economists/simple-feature-geometry-simple-feature-geometry-list-column-and-simple-feature.html)
from
[R as GIS for Economists](https://tmieno2.github.io/R-as-GIS-for-Economists/).

This uses
[sf](https://r-spatial.github.io/sf/)
and
[httr](https://httr.r-lib.org/)
R libraries, as well as
[tidyverse](https://www.tidyverse.org/).

This Rmarkdown file assumes that 
[nativeLandSetup.Rmd](https://github.com/byandell-Tribal/landMaps/blob/main/nativeLandSetup.Rmd)
has been used to set up files.
In there, the
[native-land.ca API key](https://native-land.ca/resources/api-docs/)
is stored.
Below, calls to `readRDS("data/*.rds)` bring up saved files.

```{r}
nativeLandKey <- readRDS("data/nativeLandKey.rds")
```

Here is the complete Native Land dataset, and a US subset.
This is for internal use.

```{r}
nativeLandData <- readRDS("data/NativeLandData.rds")
nativeLandUS <- readRDS("data/NativeLandUS.rds")
```

### Native Land Slug (lookup names)

The `nativeLandSlug` only has `category` and `Slug`,
useful for name lookups later. 
It would be nice if `native-land.ca` had this directly.

```{r}
nativeLandSlug <- readRDS("data/NativeLandSlug.rds")
```

This file is saved locally, and used in the shiny app
[landMaps::nativeLandApp()](https://github.com/byandell-Tribal/landMaps/blob/main/R/nativeLand.R).
However, it is not saved to github.
As constructed, it include only the `category` and `Slug`
columns from
[native-land.ca](https://native-land.ca/resources/api-docs/).

### Query for particlar territory, language and treaty combinations

This pulls all entries involving `akota` (Lakota and Nakota languages),
`sioux` (Oceti Sakowin Sioux) and `laramie` (Fort Laramie treaties).

```{r}
dplyr::filter(nativeLandSlug, grepl("akota|sioux|laramie", Slug))
```

### Plot Oceti Sakowin and Lakota Lands

This plots the Oceti Sakowin territories, languages and treaties
maps from `native-land.ca`, either together or on separate facets.

```{r}
str(nativeLandData)
```

```{r}
oceti_sakowin <- dplyr::filter(nativeLandData,
                               grepl("akota|oceti|laramie",
                                     .data$Slug))
```

```{r warning = FALSE}
landMaps::ggplot_sf(legend.position = "right") + 
  landMaps::ggplot_nativeLand(oceti_sakowin,
                              title = "Oceti Sakowin")
```

```{r}
landMaps::ggplot_sf() + 
  # Base map from OpenStreetMap
  ggspatial::annotation_map_tile(type = "osm", zoom = 14,
                                 progress = "none") +
  landMaps::ggplot_nativeLand(oceti_sakowin,
                              title = "Oceti Sakowin")
```

```{r}
landMaps::ggplot_sf() + 
  landMaps::ggplot_nativeLand(oceti_sakowin) +
  ggplot2::facet_wrap(~ category)
```

### Menominee Lands

```{r}
menominee <- dplyr::filter(nativeLandData,
                           grepl("menominee", Slug))
```

```{r warning=FALSE}
landMaps::ggplot_sf() +
  landMaps::ggplot_nativeLand(menominee)
```

The following does not work yet. Not sure why.

```{r warning=FALSE, eval=FALSE}
landMaps::ggplot_sf() +
  # Base map from OpenStreetMap
  ggspatial::annotation_map_tile(type = "osm", zoom = 14,
                                 progress = "none") +
  landMaps::ggplot_nativeLand(menominee)
```

### Direct request

The `get_nativeLand` uses the `native-land.ca` API along with the key.
The challenge is you need to know the Slug exactly,
or use a lookup data frame such as `nativeLandSlug`.

```{r}
landMaps::get_nativeLand("territories", "oceti", nativeLandKey)
```

```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps::ggplot_nativeLand(
    landMaps::get_nativeLand("territories", "oceti-sakowin-sioux",
                             nativeLandKey))
```

```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps::ggplot_nativeLand(
    landMaps::get_nativeLand("territories", "oceti",
                             nativeLandKey, nativeLandSlug))
```

## Intersections with Minnesota, Wisconsin and Michigan

```{r}
census_geometry <- readRDS("data/census_geometry.rds")
```

```{r}
census_mnwimi <- dplyr::filter(
  census_geometry,
  geography == "states",
  NAME %in% c("Wisconsin", "Minnesota", "Michigan"))
```

```{r}
bbox_mnwimi <- sf::st_bbox(
  dplyr::filter(census_geometry, geography == "states",
                NAME %in% c("Wisconsin", "Minnesota", "Michigan")))
bbox_mnwimi_sf <- tibble::tibble(x = bbox_wi[c("xmin","xmax")],
                     y = bbox_wi[c("ymin","ymax")]) |> 
  sf::st_as_sf(coords = c("x", "y")) |> 
  sf::st_set_crs(sf::st_crs(nativeLandData)) |>
  #get the bounding box of the two x & y coordintates, make sfc
  sf::st_bbox() |>
  sf::st_as_sfc()
```

```{r}
ggplot2::ggplot() +
  ggplot2::geom_sf(data = bbox_mnwimi_sf)
```

```{r warning=FALSE}
nl_bbox_mnwimi <- list()
for(i in seq_len(nrow(nativeLandData))) {
  nl_bbox_mnwimi[[i]] <- tryCatch(
    sf::st_intersection(nativeLandData[i,], bbox_mnwimi_sf),
    error = function(e) NA)
}
nl_bbox_mnwimi <- dplyr::bind_rows(
  nl_bbox_mnwimi[!is.na(nl_bbox_mnwimi)])
table(nl_bbox_mnwimi$category)
```

```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps::ggplot_nativeLand(nl_bbox_mnwimi,
                              title = "Native Land in MN/WI/MI") +
  landMaps:::ggplot_layer_sf(dplyr::filter(census_geometry, geography == "states", NAME %in% c("Wisconsin", "Minnesota", "Michigan")), color = "black") +
  ggplot2::facet_wrap(~ category)
```

```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps::ggplot_nativeLand(
    dplyr::filter(nl_bbox_mnwimi, category == "territories"),
    title = "Native Land in MN/WI/MI") +
  landMaps:::ggplot_layer_sf(dplyr::filter(census_geometry, geography == "states", NAME %in% c("Wisconsin", "Minnesota", "Michigan")), color = "black")
```

```{r}
nativeLandUS <- readRDS("data/NativeLandUS.rds")
```

```{r}
ggplot_sf(color = unique(nativeLandUS$color)) + 
  ggplot_nativeLand(dplyr::filter(nativeLandUS, category == "territories"),
                                  label = FALSE)
```

## Working stuff

```{r warning=FALSE}
nl_mnwimi <- list()
for(i in seq_len(nrow(nativeLandData))) {
  nl_mnwimi[[i]] <- tryCatch(
    sf::st_intersection(nativeLandData[i,], census_mnwimi),
    error = function(e) NA)
}
nl_mnwimi <- dplyr::bind_rows(nl_mnwimi[!is.na(nl_mnwimi)])
table(nl_mnwimi$category)
```

```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps::ggplot_nativeLand(nl_mnwimi,
                              title = "Native Land in MN/WI/MI") +
  landMaps:::ggplot_layer_sf(dplyr::filter(census_geometry, geography == "states", NAME %in% c("Wisconsin", "Minnesota", "Michigan")), color = "red") +
  ggplot2::facet_wrap(~ category)
```


```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps::ggplot_nativeLand(nl_mnwimi) +
  landMaps:::ggplot_layer_sf(dplyr::filter(census_geometry, geography == "states", NAME %in% c("Wisconsin", "Minnesota", "Michigan")), color = "red") +
  ggplot2::facet_wrap(~ category)
```

The list below was done by hand using `native-land.ca`.
There should be a programming way to find the overlap.

```{r}
wiTribes <- c("potawatomi", "brothertown", "menominee", "anishin", "ho-chunk", "oceti", "odawa", "mississauga", "fox", "wyandot",
              "myaamia", "niuachi", "dewakan", "wahpe", "yankton",
              "sisseton")
name <- dplyr::filter(
      nativeLandSlug,
      grepl(paste(wiTribes, collapse = "|"), .data$Slug),
      grepl("territories", .data$category))$Slug
nativeLandWIs <- dplyr::filter(nativeLandData,
                  .data$category == "territories",
                  .data$Slug %in% name)
nativeLandWIs$Name
```

```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps:::ggplot_layer_sf(dplyr::filter(census_geometry, geography == "states", NAME %in% c("Wisconsin", "Minnesota", "Michigan"))) +
  landMaps::ggplot_nativeLand(nativeLandWIs,
                              title = "MN/WI/MI embedded in Native Lands")
```

```{r warning = FALSE}
landMaps::ggplot_sf() +
  landMaps:::ggplot_layer_sf(dplyr::filter(census_geometry, geography == "states", NAME %in% c("Wisconsin", "Minnesota", "Michigan"))) +
  landMaps::ggplot_nativeLand(nativeLandWIs,
                              title = "Native Lands overlapping with MN/WI/MI") +
  ggplot2::coord_sf(xlim = bbox_wi[c("xmin","xmax")],
                    ylim = bbox_wi[c("ymin","ymax")])

```
