---
title: "NativeLand"
author: "Brian Yandell"
date: "2024-10-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = ".")
```

## Native Land Local Data Setup

This Rmarkdown file assumes that 
[censusSetup.Rmd](https://github.com/byandell-envsys/landMaps/blob/main/censusSetup.Rmd)
has been already used to set up US Census files.

The
[native-land.ca](https://native-land.ca/resources/api-docs/)
just reorganized their API.
I have put my API key in folder `data` local to my computer.
You can cut and paste your own

```{r eval=FALSE}
if(!dir.exists("data") dir.create("data"))
saveRDS(readline(), "data/nativeLandKey.rds")
```

```{r}
nativeLandKey <- readRDS("data/nativeLandKey.rds")
```

### Native Land Full Downloads

Get full downloads at <https://api-docs.native-land.ca/full-geojsons>
using your own key.
I have done that and put entries in my (local) `data` folder.
Here I read those GeoJSON files into a `sf` data frame.

I save this locally for my own possible use.
This object is **NOT** used in the package, nor is it
saved to
[github](https://github.com/byandell-envsys/nativeLand).

The following reads all Native land features and
organizes them into `category`.

```{r}
if(file.exists(filename <- "data/NativeLandData.rds")) {
  nativeLandData <- readRDS(filename)
} else {
  nativeLandData <- dplyr::bind_rows(
    territories = landMaps::features_reform(
      rjson::fromJSON(file = "data/territories")$features),
    languages   = landMaps::features_reform(
      rjson::fromJSON(file = "data/languages")$features),
    treaties    = landMaps::features_reform(
      rjson::fromJSON(file = "data/treaties")$features),
    .id = "category")
  sf::st_crs(nativeLandData) <- 4326
  saveRDS(nativeLandData, filename)
}
```

### Native Land Slug (lookup names)

The `nativeLandSlug` only has `category` and `Slug`,
useful for name lookups later. 
It would be nice if `native-land.ca` had this directly.

```{r}
if(file.exists(filename <- "data/NativeLandSlug.rds")) {
  nativeLandSlug <- readRDS(filename)
} else {
  nativeLandSlug <- dplyr::select(nativeLandData, category, Slug)
  nativeLandSlug$geometry <- NULL
  saveRDS(nativeLandSlug, filename)
}
```

This `Slug` file is saved locally, and used in the shiny app
[landMaps::nativeLandApp()](https://github.com/byandell-envsys/landMaps/blob/main/R/nativeLand.R).
However, it is not saved to github.
As constructed, it include only the `category` and `Slug`
columns from
[native-land.ca](https://native-land.ca/resources/api-docs/).

### nativeLand for US only.

```{r}
census_geometry <- readRDS("data/census_geometry.rds")
```
.
```{r}
if(file.exists(filename <- "data/NativeLandUS.rds")) {
  nativeLandUS <- readRDS(filename)
} else {
  nativeLandUS <- landMaps:::nativeLand_us(nativeLandData, census_geometry)
  saveRDS(nativeLandUS, "data/nativeLandUS.rds")
}
```
